---
title: "LME"
author: "Claudia Morales Valiente"
date: "2024-09-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prerequisites

The following R libraries are required for the code to run:

- `dplyr`: Data manipulation
- `janitor`: Data cleaning
- `lme4`: Fitting Linear and Generalized Linear Mixed-Effects Models
- `lmerTest`: Tests in linear mixed-effects models
- `emmeans`: Estimated marginal means and pairwise comparisons
- `effects`: Visualizing interaction effects
- `sjPlot`: Plotting tools for social science data

## Code Breakdown

### Data Loading and Cleaning

The dataset (`file_name.csv`) is loaded, and the variable names are cleaned using `janitor::clean_names()` for consistency.

```{r database}
db <- read_csv("file_name.csv")
db <- clean_names(db)
```

### Variable Definitions

The code defines key variables for the analysis:

- **Random effects**: `random1` and `random2` (factors for random variables)
- **Predictors**: `predictor1` and `predictor2`
- **Outcome**: `outcome`

```{r variables}
random1 <- factor(db$random_variable1)
random2 <- factor(db$random_variable2)
predictor1 <- factor(db$predictor1)
predictor2 <- factor(db$predictor2)
outcome <- db$outcome
```

### Checking for Missing Values

The code checks for missing values in the dataset row-wise and column-wise and prints messages accordingly.

```{r NA}
any_na <- any(is.na(db))
any_na_row <- apply(db, 1, function(row) any(is.na(row)))
any_na_col <- apply(db, 2, function(col) any(is.na(col)))

if (any_na) {
  print("There are NA values in the matrix.")
} else {
  print("There are no NA values in the matrix.")
}

if (any(any_na_row)) {
  print("There are NA values in one or more rows of the matrix.")
} else {
  print("There are no NA values in any rows of the matrix.")
}

if (any(any_na_col)) {
  print("There are NA values in one or more columns of the matrix.")
} else {
  print("There are no NA values in any columns of the matrix.")
}
```

### Model Building

The following models are fitted using linear mixed models (`lmer`):

- **Model 1a**: Null model with `random1` as a random effect
- **Model 1b**: Null model with `random2` as a random effect
- **Model 2**: Main effect of `predictor1` with `random1` and `random2` as random effects
- **Model 3**: Main effect of `predictor2` with `random1` and `random2` as random effects
- **Model 4**: Main effects of both `predictor1` and `predictor2` with `random1` and `random2`
- **Model 5**: Interaction between `predictor1` and `predictor2` with `random1` and `random2`

```{r models}
model1a <- lmer(outcome ~ 1 + (1|random1), data = db, REML = FALSE)
model1b <- lmer(outcome ~ 1 + (1|random2), data = db, REML = FALSE)

model2 <- lmer(outcome ~ predictor1 + (1|random1) + (1|random2), data = db, REML = FALSE)
model3 <- lmer(outcome ~ predictor2 + (1|random1) + (1|random2), data = db, REML = FALSE)
model4 <- lmer(outcome ~ predictor1 + predictor2 + (1|random1) + (1|random2), data = db, REML = FALSE)
model5 <- lmer(outcome ~ predictor1 * predictor2 + (1|random1) + (1|random2), data = db, REML = FALSE)
```

### Model Comparisons

The `anova` function is used to compare models to determine the effects of the predictors.

```{r comparisons}
anova(model4, model3) # Main effect of predictor1
anova(model4, model2) # Main effect of predictor2
anova(model5, model4) # Interaction effect
```

### Post-Hoc Comparisons (in case of interaction)

If an interaction between `predictor1` and `predictor2` is found, post-hoc comparisons are conducted using the `emmeans` package. The results are then converted into a data frame, and significance is indicated.

```{r pairwise comparisons}
emm <- emmeans(model5, ~ predictor1 * predictor2)
pairwise_comparisons <- pairs(emm)
adjusted_comparisons <- summary(pairwise_comparisons)
results_df <- as.data.frame(adjusted_comparisons)
results_df$Significant <- results_df$p.value < 0.05
print(results_df)
```

## Usage

1. Ensure the dataset is saved as `file_name.csv` in the appropriate directory.
2. Install the required libraries.
3. Run the script to fit the models and generate results.
4. The script will print model comparison results and indicate significant post-hoc comparisons, if any.
